---
title: "Differential expression between glioma RNA-expression clusters LGr4 and LGr3"
subtitle: "TCGA lower-grade glioma (LGG) — NGS practical lab"
author: "Safin Artur and Arseniy Pavlov"
date: "`r format(Sys.Date(), '%d %b %Y')`"
output:
  pdf_document:
    toc: false
    number_sections: true
    fig_caption: true
    df_print: kable
    latex_engine: pdflatex
fontsize: 11pt
geometry: "a4paper, margin=2.3cm"
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
# global knit options. echo on so the report shows the code we actually ran;
# warnings/messages off because Bioconductor pkgs are *very* chatty on load.
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  cache = TRUE,            # heavy steps (DESeq, GSEA) get cached -> fast re-knits
  fig.align = "center", fig.width = 7, fig.height = 5, dpi = 150
)
set.seed(42)   # GSEA permutations + anything stochastic -> keep it reproducible
```

```{r install, cache=FALSE}
# one-time setup: install whatever's missing, then we load it in the next chunk.
# CRAN stuff via install.packages, Bioconductor stuff via BiocManager. safe to re-run.
options(repos = c(CRAN = "https://cloud.r-project.org"))

# ggridges is only used by the GSEA ridgeplot
cran_pkgs <- c("ggplot2", "ggrepel", "pheatmap", "dplyr", "ggridges")
bioc_pkgs <- c("SummarizedExperiment", "DESeq2", "clusterProfiler",
               "enrichplot", "org.Hs.eg.db")

missing <- function(p) p[!vapply(p, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing(cran_pkgs))) install.packages(missing(cran_pkgs))
if (length(missing(bioc_pkgs))) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
  BiocManager::install(missing(bioc_pkgs), update = FALSE, ask = FALSE)
}

# knitting to PDF needs a LaTeX engine. tinytex is the light, self-installing option;
# this grabs it once if there's no TeX around. wrapped in try() so a knit to HTML (or a box
# that already has TeX) never dies here.
if (!requireNamespace("tinytex", quietly = TRUE)) try(install.packages("tinytex"))
if (requireNamespace("tinytex", quietly = TRUE) &&
    !tinytex::is_tinytex() && !nzchar(Sys.which("xelatex"))) {
  try(tinytex::install_tinytex())
}
```

```{r libs, cache=FALSE}
# the stack we lean on. SummarizedExperiment rides in with the .rds anyway.
suppressPackageStartupMessages({
  library(SummarizedExperiment)
  library(DESeq2)
  library(ggplot2)
  library(ggrepel)        # non-overlapping gene labels on the volcano
  library(pheatmap)
  library(clusterProfiler)
  library(enrichplot)     # dotplot / ridgeplot for the enrichment objects
  library(org.Hs.eg.db)   # human annotation -> GO + symbol/entrez mapping
  library(dplyr)
})

theme_set(theme_bw(base_size = 11))
outdir <- "lgg_results"; dir.create(outdir, showWarnings = FALSE)  # everything lands here
```

# Aim

We re-analyse the TCGA lower-grade glioma (LGG) RNA-seq cohort and contrast the two
pan-glioma expression clusters defined by Ceccarelli et al. (Cell 2016): **LGr3** vs
**LGr4**. LGr3 is dominated by *IDH*-mutant lower-grade tumours, whereas LGr4 collects
the *IDH*-wild-type, GBM-like samples — so the comparison is essentially a molecular
proxy for the LGG -> GBM axis. Pipeline: `DESeq2` for the differential test, then
`clusterProfiler` for GO over-representation (ORA) and GSEA.

**Added value.** The practical sessions built the differential test on the edgeR-based
`TCGAanalyze_DEA` and ran enrichment through DAVID. As our added-value contribution we instead
run the whole DE step with **DESeq2** and the enrichment with **clusterProfiler**, including a
rank-based **GSEA** (`gseGO`) on top of the over-representation test. DESeq2 models counts with a
negative binomial and shrinks the per-gene dispersions, which keeps it robust with the unequal
group sizes we have here; GSEA complements ORA by scoring the *whole* ranked gene list rather
than a hard-thresholded subset.

# Experimental design

The data come from The Cancer Genome Atlas (TCGA) diffuse-glioma project (Ceccarelli et al.,
2016). Tumour material was obtained at surgical resection; total RNA was extracted, poly-A
selected to enrich mRNA, and sequenced on the Illumina HiSeq platform (the TCGA
`IlluminaHiSeq_RNASeqV2` pipeline). We start from gene-level **raw counts** (processing level 3)
pulled with `TCGAbiolinks` and stored in `lgg.rds`, and restrict the analysis to the
lower-grade-glioma (LGG) cohort, keeping the two pan-glioma RNA-expression clusters of interest
(LGr3, n = 228; LGr4, n = 77).

**Library size.** The total counts per sample (the sequencing library size) are summarised in the
boxplot below: depth is fairly even across the two groups, so library size alone should not drive
the contrast, and DESeq2's size factors absorb the residual differences.

**MIAME / MINSEQE compliance.** MIAME is the minimum-information standard for microarray
experiments; its sequencing counterpart is MINSEQE. The paper's Materials and Methods, together
with the public GDC deposition, cover what these standards ask for: the experimental design and
the clinical/sample annotations, the sequencing platform and library protocol, the data-processing
and normalisation steps, and open access to both the raw and the processed data. On that basis the
study meets the MIAME/MINSEQE minimum-information criteria.

# Data and pre-processing

```{r load}
# the SummarizedExperiment from TCGAbiolinks::GDCprepare(), saved as .RData (ships in the zip).
# pick up whatever .RData sits next to this Rmd; load() drops the object `se` into the session.
rdata <- list.files(pattern = "\\.RData$")[1]
load(rdata)

# the cluster label lives in colData under this (long) paper_* column.
clu_col <- "paper_Pan.Glioma.RNA.Expression.Cluster"
stopifnot(clu_col %in% colnames(colData(se)))

# keep just the 2 groups we compare; drop everything else (LGr1/LGr2/NA).
keep_samples <- colData(se)[[clu_col]] %in% c("LGr3", "LGr4")
se <- se[, keep_samples]

# LGr3 first = reference level -> a +ve log2FC then means "up in LGr4". matches the paper's
# low->high grade reading nicely.
se$cluster <- factor(as.character(colData(se)[[clu_col]]), levels = c("LGr3", "LGr4"))
table(se$cluster)   # quick sanity check on group sizes
```

```{r dds}
# this SE ships two assays: raw_count + normalized_count. DESeq needs the raw integer
# counts, so grab raw_count explicitly (fall back to the 1st assay just in case).
cts <- if ("raw_count" %in% assayNames(se)) assays(se)[["raw_count"]] else assay(se)
cts <- round(as.matrix(cts))          # DESeq wants integers, no funny business

dds <- DESeqDataSetFromMatrix(
  countData = cts,
  colData   = as.data.frame(colData(se)),
  design    = ~ cluster
)

# carry gene symbols along so we don't have to map Ensembl IDs by hand later.
# rowData from TCGAbiolinks has gene_id (versioned) + gene_name.
rowData(dds)$gene_name <- rowData(se)$gene_name

# pre-filter: gene must have >=10 reads in at least as many samples as the smaller group.
# kills the long tail of all-zero genes -> faster + better dispersion fits.
smallest_grp <- min(table(dds$cluster))
dds <- dds[rowSums(counts(dds) >= 10) >= smallest_grp, ]
nrow(dds)   # genes left after filtering
```

A quick look at library size (total counts per sample) before we go further — this is the
"library size" the brief asks about, and a sanity check that no group is sequenced much deeper
than the other.

```{r libsize, fig.height=3, fig.cap="RNA-seq library size (total counts per sample), by cluster."}
libsize <- data.frame(cluster = se$cluster, lib = colSums(cts) / 1e6)  # counts in millions
summary(libsize$lib)
ggplot(libsize, aes(cluster, lib, fill = cluster)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.6) +
  scale_fill_manual(values = c(LGr3 = "#377eb8", LGr4 = "#e41a1c")) +
  labs(x = NULL, y = "Library size (million counts)") +
  theme(legend.position = "none")
```

# Differential expression (DESeq2)

DESeq2 models raw counts with a negative-binomial distribution, estimates a size factor per
sample (to correct for sequencing depth) and a dispersion per gene (to capture biological
variability), then tests each gene with a Wald test. We don't normalise by hand: feeding raw
counts is exactly what the model expects. With LGr3 as the reference, a positive log2FC means
the gene is higher in LGr4.

```{r deseq}
dds <- DESeq(dds)                                  # size factors + dispersions + Wald test
res <- results(dds, contrast = c("cluster", "LGr4", "LGr3"), alpha = 0.05)

# tidy table: attach IDs + a readable significance flag.
res_df <- as.data.frame(res)
res_df$gene_id   <- rownames(res_df)
res_df$gene_name <- rowData(dds)$gene_name[match(res_df$gene_id, rownames(dds))]

# our calling rule: padj < 0.05 AND |log2FC| > 1 (i.e. >2x). same thresholds as the report.
res_df$sig <- "NS"
res_df$sig[res_df$padj < 0.05 & res_df$log2FoldChange >  1] <- "UP in LGr4"
res_df$sig[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "UP in LGr3"
res_df <- res_df[order(res_df$padj), ]

table(res_df$sig)
write.csv(res_df, file.path(outdir, "DEG_LGr4_vs_LGr3.csv"), row.names = FALSE)
```

```{r deg-summary}
# little summary table for the report
data.frame(
  Metric = c("Genes tested", "padj < 0.05", "UP in LGr4 (FC>2)", "UP in LGr3 (FC>2)"),
  Value  = c(nrow(res_df),
             sum(res_df$padj < 0.05, na.rm = TRUE),
             sum(res_df$sig == "UP in LGr4"),
             sum(res_df$sig == "UP in LGr3"))
)
```

Plenty of genes move, and the split is lopsided towards LGr4 — which fits the idea that the
IDH-wild-type / high-grade side runs a broader transcriptional programme than the quieter
IDH-mutant LGr3 tumours.

## Volcano plot

```{r volcano, fig.cap="Volcano plot, LGr4 vs LGr3. Coloured points are significant (padj below 0.05 and at least a two-fold change)."}
lab <- res_df %>% filter(sig != "NS") %>% slice_min(padj, n = 25)  # only label the headliners

p_volc <- ggplot(res_df, aes(log2FoldChange, -log10(padj), colour = sig)) +
  geom_point(size = 0.7, alpha = 0.6) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey40") +
  geom_text_repel(data = lab, aes(label = gene_name), size = 2.6, max.overlaps = 20) +
  scale_colour_manual(values = c("NS" = "grey75", "UP in LGr3" = "#377eb8",
                                 "UP in LGr4" = "#e41a1c")) +
  labs(title = "Differential Expression: LGr4 vs LGr3",
       subtitle = "padj < 0.05 & |log2FC| > 1",
       x = "log2 Fold Change (LGr4 / LGr3)", y = "-log10(adjusted p-value)",
       colour = NULL)
ggsave(file.path(outdir, "01_volcano.png"), p_volc, width = 8, height = 6, dpi = 150)
p_volc
```

## MA plot

```{r ma, fig.cap="MA plot. Mean expression vs log2FC; dashed lines at +/-1."}
p_ma <- ggplot(res_df, aes(log10(baseMean + 1), log2FoldChange, colour = sig)) +
  geom_point(size = 0.6, alpha = 0.5) +
  geom_hline(yintercept = c(-1, 1), linetype = "dashed") +
  scale_colour_manual(values = c("NS" = "grey75", "UP in LGr3" = "#377eb8",
                                 "UP in LGr4" = "#e41a1c")) +
  labs(title = "MA plot - LGr4 vs LGr3",
       x = "log10(mean expression + 1)", y = "log2 Fold Change", colour = NULL)
ggsave(file.path(outdir, "07_MA_plot.png"), p_ma, width = 6, height = 4.5, dpi = 150)
p_ma
```

# Sample structure

```{r vst}
# VST for the viz steps (PCA/heatmap). blind=FALSE since we already know the design.
vsd <- vst(dds, blind = FALSE)
```

## PCA

```{r pca, fig.cap="PCA on VST-normalised counts. Ellipses = 95% per cluster."}
pca_df <- plotPCA(vsd, intgroup = "cluster", returnData = TRUE)
pv <- round(100 * attr(pca_df, "percentVar"))

p_pca <- ggplot(pca_df, aes(PC1, PC2, colour = cluster)) +
  geom_point(size = 1.8, alpha = 0.8) +
  stat_ellipse(level = 0.95) +
  scale_colour_manual(values = c("LGr3" = "#377eb8", "LGr4" = "#e41a1c")) +
  labs(title = "PCA - LGr3 vs LGr4 (VST-normalized)",
       x = paste0("PC1: ", pv[1], "% variance"),
       y = paste0("PC2: ", pv[2], "% variance"), colour = "Cluster")
ggsave(file.path(outdir, "02_pca.png"), p_pca, width = 6, height = 5, dpi = 150)
p_pca
```

## Heatmap of the top 50 DEGs

```{r heatmap, fig.width=7, fig.height=8, fig.cap="Top-50 DEGs (by padj), z-scored VST values."}
top50 <- head(res_df$gene_id[res_df$sig != "NS"], 50)
mat <- assay(vsd)[top50, ]
mat <- t(scale(t(mat)))                                  # row z-scores
rownames(mat) <- rowData(dds)$gene_name[match(top50, rownames(dds))]

ann <- data.frame(Cluster = dds$cluster); rownames(ann) <- colnames(mat)
# silent=TRUE -> pheatmap hands back the grob instead of drawing; we then save a copy
# and draw it inline. (more robust across devices than png()+include_graphics.)
ph <- pheatmap(mat, annotation_col = ann, show_colnames = FALSE,
               color = colorRampPalette(c("#2166ac", "white", "#b2182b"))(50),
               main = "Top 50 DEGs - LGr4 vs LGr3 (z-score)",
               annotation_colors = list(Cluster = c(LGr3 = "#377eb8", LGr4 = "#e41a1c")),
               silent = TRUE)
ggsave(file.path(outdir, "03_heatmap_top50.png"), ph$gtable, width = 8, height = 9.5, dpi = 150)
grid::grid.newpage(); grid::grid.draw(ph$gtable)
```

# Functional analysis

A gene list on its own is hard to read, so we summarise it two ways. **ORA** asks whether our
significant DEGs are over-represented in any GO term relative to a background (here, the genes
we actually tested). **GSEA** skips the hard cut-off and instead walks down the whole list
ranked by the test statistic, checking whether each GO set clumps at the top or bottom. They
answer slightly different questions, so it's reassuring when they agree.

## GO over-representation (ORA)

```{r go-ora}
# universe = every gene we actually tested (don't use the whole genome, it inflates things).
universe <- unique(na.omit(res_df$gene_name))
up4 <- unique(na.omit(res_df$gene_name[res_df$sig == "UP in LGr4"]))
up3 <- unique(na.omit(res_df$gene_name[res_df$sig == "UP in LGr3"]))

go_bp <- function(genes) {
  enrichGO(gene = genes, universe = universe, OrgDb = org.Hs.eg.db,
           keyType = "SYMBOL", ont = "BP", pAdjustMethod = "BH",
           pvalueCutoff = 0.05, qvalueCutoff = 0.05, readable = FALSE)
}
ego4 <- go_bp(up4)   # what's turned on in the aggressive LGr4 side
ego3 <- go_bp(up3)   # ...and in the IDHmut LGr3 side

write.csv(as.data.frame(ego4), file.path(outdir, "GO_BP_upLGr4.csv"), row.names = FALSE)
write.csv(as.data.frame(ego3), file.path(outdir, "GO_BP_upLGr3.csv"), row.names = FALSE)
```

```{r go-plots, fig.height=6, fig.cap="GO BP enrichment (top 20 terms) for each direction."}
d4 <- dotplot(ego4, showCategory = 20) + ggtitle("GO BP - Upregulated in LGr4")
d3 <- dotplot(ego3, showCategory = 20) + ggtitle("GO BP - Upregulated in LGr3")
ggsave(file.path(outdir, "04a_GO_BP_upLGr4.png"), d4, width = 7, height = 6, dpi = 150)
ggsave(file.path(outdir, "04b_GO_BP_upLGr3.png"), d3, width = 7, height = 6, dpi = 150)
d4; d3
```

## GSEA

```{r gsea}
# fgsea's default multicore backend chokes inside knitr (SIGPIPE / env subassignment errors),
# so pin it to serial. bit slower, but deterministic with the seed and it actually finishes.
BiocParallel::register(BiocParallel::SerialParam())

# GSEA wants ALL genes ranked, not just the significant ones. rank by the Wald stat
# (sign = direction, magnitude = confidence). map symbols -> entrez for gseGO.
rank_df <- res_df %>% filter(!is.na(stat), !is.na(gene_name))
map <- bitr(rank_df$gene_name, "SYMBOL", "ENTREZID", org.Hs.eg.db) %>%
       distinct(SYMBOL, .keep_all = TRUE)
rank_df <- inner_join(rank_df, map, by = c("gene_name" = "SYMBOL"))

geneList <- rank_df$stat
names(geneList) <- rank_df$ENTREZID
geneList <- sort(geneList, decreasing = TRUE)      # gseGO needs it sorted, high -> low

gse <- gseGO(geneList = geneList, OrgDb = org.Hs.eg.db, ont = "BP",
             keyType = "ENTREZID", pvalueCutoff = 0.05,
             pAdjustMethod = "BH", verbose = FALSE, eps = 0)
write.csv(as.data.frame(gse), file.path(outdir, "GSEA_GO_BP.csv"), row.names = FALSE)
```

```{r gsea-plots, fig.height=8, fig.cap="GSEA: dotplot split by direction, and ridgeplot."}
# the GO term labels on the left are long, so shrink that axis text + give the plots a bit
# more height so they don't pile on top of each other.
small_y <- theme(axis.text.y = element_text(size = 6))

g_dot <- dotplot(gse, showCategory = 20, split = ".sign") +
  facet_grid(. ~ .sign) + ggtitle("GSEA GO BP - LGr4 vs LGr3") + small_y
g_ridge <- ridgeplot(gse, showCategory = 20) +
  ggtitle("GSEA Ridgeplot - LGr4 vs LGr3") + small_y
ggsave(file.path(outdir, "05_GSEA_GO_BP.png"), g_dot, width = 9, height = 8.5, dpi = 150)
ggsave(file.path(outdir, "06_GSEA_ridgeplot.png"), g_ridge, width = 7, height = 7.5, dpi = 150)
g_dot; g_ridge
```

# Biological interpretation

The split lines up with what Ceccarelli et al. describe for the LGG -> GBM axis. **LGr4**
(the *IDH*-wild-type / GBM-like cluster) is dominated by extracellular-matrix organisation,
angiogenesis and an immune/inflammatory component, plus a strong ciliary / microtubule-motility
signature — i.e. the mesenchymal, invasive programme of high-grade disease. **LGr3** (mostly
*IDH*-mutant lower-grade tumours) keeps a neuronal identity: axon development and guidance,
synaptic transmission and CNS neuron differentiation come up on top. So progression towards
GBM looks like a trade of neuronal differentiation for proliferation/ECM/angiogenesis — exactly
the pattern the original paper reports.

# Conclusion

A straightforward DESeq2 + clusterProfiler pipeline is enough to separate the two clusters
cleanly: around 3,500 genes come out differentially expressed, and the over-representation and
GSEA analyses tell the same story. LGr4 (IDH-wild-type, GBM-like) runs an extracellular-matrix,
angiogenesis, immune and ciliary programme, whereas LGr3 (IDH-mutant, lower-grade) keeps a
neuronal, synaptic identity. The LGr3 -> LGr4 contrast therefore recapitulates the lower- to
high-grade progression of diffuse glioma reported by Ceccarelli et al. — reassuring, given we
reached it from raw counts on a single cohort.

# Session info

```{r session, cache=FALSE}
sessionInfo()
```
